# Hypothesis tests and measures of association (IES Week 9) {#sec-09-comparison}

This practical follows the IES Week 9 seminar of the same name. That seminar set out the logic of a hypothesis test: stating a null and an alternative hypothesis, turning a test statistic into a p-value, and being precise about what a p-value is and what it is not. It also showed that a confidence interval and a hypothesis test carry the same information, viewed two ways, and it established the reporting standard we use from here on: an effect size, a 95% confidence interval, and a p-value, reported together. Finally, it reactivated the risk ratio and odds ratio from the epidemiology half of the programme, attaching a confidence interval and a p-value to numbers you can already calculate by hand. In this session we put all of that into R.

## Chapter intended learning outcomes

By the end of this session you will be able to:

1. Perform and interpret an independent-samples t-test in R
2. Perform and interpret a paired t-test, recognise from a data structure which is required, and distinguish paired measurements taken over time from two related measurements taken at once
3. Construct a contingency table and perform a chi-squared test
4. Obtain risk ratios and odds ratios with confidence intervals, and reconcile them with the hand calculation taught in the epidemiology half
5. Write up a result in the style expected in a public health journal

## Setup

As in IES Weeks 7 and 8, we read the data with `read_csv()` and turn the six coded variables into factors against the dictionary. We also bring back `strep_tb`, from the `medicaldata` package, for the categorical half of this chapter.

This chapter adds one new package, `epitools`. It isn't part of the tidyverse, and it won't already be installed alongside the packages we've used so far. As IES Week 4 explained, installing a package and loading it are two different actions. If `epitools` isn't already on your machine, run `install.packages("epitools")` once in the console, not inside this document. Once it's installed, `library(epitools)` below makes its functions available for this session, exactly like every other package we load.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(here)
library(medicaldata)
library(epitools)

nhanes <- read_csv(here("data", "mph_nhanes_20241107.csv"))
dictionary <- read_csv(here("data", "mph_nhanes_20241107_dictionary.csv"))
```

As in IES Week 8, `dictionary` is read in even though no code below refers to it. It's there so it's to hand the moment you're unsure what a column means.

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
source(here("R", "gtsummary-knitprint.R"))
```

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c("Male", "Female")),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c("Mexican American", "Other Hispanic", "White", "Black", "Others")),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c("Less than high school", "High school graduate", "Above high school")),
    smoking = factor(smoking,
      levels = c("Never", "Previous", "Current")),
    phq.c = factor(phq.c,
      levels = c("0 Normal", "1 Mild symptoms", "2 Possible depression"),
      labels = c("Normal", "Mild symptoms", "Possible depression"),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c("No", "Yes"))
  )
```

This is the same recode we introduced in IES Week 7 and repeated in IES Week 8. See IES Week 7 for the reasoning behind each variable and each choice of `ordered = TRUE`. We repeat it here, unchanged, so this chapter stands on its own too.

## Independent-samples t-test: cholesterol by gender

`t.test()` compares two independent groups directly. We already used it in IES Week 8 to build a confidence interval for the difference in total cholesterol (`chol`, in mmol/L) between men and women:

```{r}
#| label: ttest-chol-gender
t.test(chol ~ gender, data = nhanes |> filter(!is.na(gender)))
```

This is the exact same call we made in IES Week 8. `t.test()` has been giving us a hypothesis test the whole time. In IES Week 8, hypothesis testing hadn't been introduced yet, so we only looked at the confidence interval and set the rest of the output aside. Now we bring the rest into view: a t-statistic, degrees of freedom, and a p-value.

Before trusting a t-test, it's worth checking that the two groups aren't wildly different in spread or shape, the way IES Week 7's boxplot of `chol` by `gender` already let us do:

```{r}
#| label: boxplot-recheck
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = chol)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Total cholesterol (mmol/L)")
```

The two boxplots are similar in spread and shape, just shifted slightly. Nothing here rules out a t-test. `t.test()` also doesn't assume equal variance between the two groups by default, which protects us even where the spreads differ more than they do here.

Before reading the interval, check the direction, exactly as IES Week 8 set out. The alternative-hypothesis line says *group Male and group Female*, because `"Male"` is the first level of `gender`, so R has printed men minus women and both ends come out negative. Reversing every sign together turns it into the more natural reading: mean total cholesterol was about 0.17 mmol/L higher in women than in men (95% CI: 0.15 to 0.19 mmol/L, p < 0.001). The two statements are the same result written from opposite ends. This is the reporting standard the seminar introduced: effect size, confidence interval, and p-value, together, every time, from here on. `p < 0.001` is how we write a p-value that R prints as something like `< 2.2e-16`. Convention reports a very small p-value this way rather than typing out the full number.

## Paired t-test: two domains of activity in one person

`nhanes_mph` records physical activity as two separate totals for each participant, both in MET-hours/day: `met_w`, activity at work, and `met_r`, activity outside work. These are two different domains of activity, measured at the same time, in the same person. They are not the same activity measured twice at two different points in time.

That distinction decides which test applies. An independent-samples t-test, like the one we just ran on `chol` by `gender`, compares two separate groups of people. Here we have one group of people, each measured twice. The right tool is a paired t-test. Rather than comparing the two columns directly, it collapses them into a single column of within-person differences (`met_w` minus `met_r`) and tests whether that difference is, on average, zero.

`t.test()` runs a paired test when we set `paired = TRUE`:

```{r}
#| label: paired-ttest
t.test(nhanes$met_w, nhanes$met_r, paired = TRUE)
```

On average, work activity ran about 3.07 MET-hours/day higher than activity outside work (95% CI: 2.93 to 3.21 MET-hours/day, p < 0.001).

Framing this as two domains of activity within a person, rather than as a change over time, is deliberate. A paired t-test also arises in a before-and-after design, where the same measurement is taken at two time points on the same person. That is a genuinely different study design from the one here, even though the R code looks identical. There is no repeated measurement over time in this data, and describing the result as though there were would misdescribe the study and plant a habit that causes real trouble in a longitudinal design later on.

## A non-parametric alternative: the Wilcoxon signed-rank test

A paired t-test assumes the within-person differences are at least roughly normally distributed. Let's check that assumption directly, the way we checked spread with a boxplot earlier:

```{r}
#| label: hist-met-diff
nhanes |>
  mutate(met_diff = met_w - met_r) |>
  ggplot(aes(x = met_diff)) +
  geom_histogram(binwidth = 5) +
  labs(x = "Work minus non-work activity (MET-hours/day)", y = "Count")
```

The distribution of differences is heavily skewed, and a large share of participants sit exactly at zero. Many people report no occupational activity at all, so `met_w` and `met_r` are equal for them. This is exactly the shape a paired t-test's normality assumption doesn't cope well with.

`wilcox.test()`, from base R, tests the same null hypothesis (no systematic difference within the pair) without assuming a normal distribution. It works with the ranks of the differences rather than their average, so one extreme value or an oddly-shaped distribution can't dominate the result the way it can dominate a mean. With `paired = TRUE`, it runs the paired version, known as the Wilcoxon signed-rank test:

```{r}
#| label: wilcox-met
wilcox.test(nhanes$met_w, nhanes$met_r, paired = TRUE)
```

The conclusion doesn't change here: both tests reject the null hypothesis convincingly, each with a p-value far below 0.001. With a sample this large, the t-test and the Wilcoxon test agree despite the skew. That won't always be true. With a smaller sample, or a more extreme distribution, the two tests can disagree, and the non-parametric result is usually the one to trust when the normality assumption is in real doubt.

## Chi-squared test: `strep_tb`'s primary result

`strep_tb`'s primary outcome is binary: did chest X-ray improve at six months (`improved`), and does that depend on trial arm (`arm`)? `table()` builds a two-way count, the same way it did in IES Week 8, and `chisq.test()`, from base R, tests whether the two variables are associated:

```{r}
#| label: table-arm-improved
tab_arm <- table(strep_tb$arm, strep_tb$improved)
tab_arm
```

```{r}
#| label: chisq-arm-improved
chisq.test(tab_arm, correct = FALSE)
```

The chi-squared statistic is large and the p-value is small: the proportion improved is not the same in the two arms.

Before trusting a chi-squared test, it's worth checking the assumption behind it. Every **expected** count, not observed count, should be reasonably large, conventionally at least 5. `chisq.test()` calculates these expected counts internally and stores them in `$expected`:

```{r}
#| label: chisq-expected
chisq.test(tab_arm, correct = FALSE)$expected
```

All four expected counts are comfortably above 5, so the test is on solid ground here. When expected counts run low, typically in a small or very unevenly split sample, `fisher.test()`, also from base R, tests the same association exactly rather than by approximation, and doesn't depend on that condition.

### Why `correct = FALSE`

We passed `correct = FALSE` above, and it's worth saying what that switches off. Left to itself on a 2×2 table, `chisq.test()` applies **Yates' continuity correction**, which shrinks the test statistic slightly to compensate for using a smooth curve to approximate counts that can only be whole numbers. It's a conservative adjustment: it makes the p-value larger.

You've already seen it in action, without being told what it was. IES Week 8 ran `prop.test()` on these same two arms and its output was headed "with continuity correction", because that function applies it by default too. The p-value it printed was 0.0003548. The uncorrected test we just ran gives 0.000166. Same data, same comparison, two different numbers, and the whole difference is this one default.

That correction earns its place when the table is thin, which is the situation it was designed for: small totals, or any expected cell count down around 5 to 10, where the approximation is under real strain. Look at the expected counts we just printed. The smallest is above 25, on 107 patients split fairly evenly four ways. Nothing here is straining, so the correction adjusts for a problem this table doesn't have, and we switch it off and read the plain Pearson test instead.

There's a practical reason too, which the next section depends on. Turning the correction off is also what makes this p-value match the one `epitools` is about to report for the same table.

## Risk ratios and odds ratios, with confidence intervals

The epidemiology half already taught how to calculate a risk ratio and an odds ratio from a 2×2 table like `tab_arm` above, and what each one means. We don't re-derive those formulas here. What's new is attaching a confidence interval and a p-value to numbers you can already compute by hand, and checking that R's version agrees with yours.

`riskratio()` and `oddsratio()`, both from `epitools`, take a 2×2 table and compare its rows, treating the first row as the reference category everything else is measured against.

They expect the table a particular way round, and that's worth knowing before building one, because getting it wrong produces a perfectly ordinary-looking answer rather than an error. The **exposure goes in the rows**, with the reference group first. The **outcome goes in the columns**, and the second column is taken as the event: the thing whose risk we're comparing. Our table has `arm` in the rows and `improved` in the columns, and `improved` holds `TRUE`/`FALSE` values, which `table()` orders `FALSE` then `TRUE`, so improvement lands in the second column exactly as needed. Print the table and look at it before passing it on. If the reference row or the event column is the wrong one, these functions still return a risk ratio, just an upside-down one.

`levels()`, from base R, lists a factor's categories in their current order, which is exactly what we need to check before comparing rows:

```{r}
#| label: levels-arm
levels(strep_tb$arm)
```

`Streptomycin` is listed first, which would make it the reference, backwards from what we want here. `relevel()`, met in IES Week 7 when we checked the reference level on `eth`, moves the level named in `ref` to the front, making it the new reference level. There we were only demonstrating it; here it changes the answer:

```{r}
#| label: relevel-arm
arm_control_ref <- relevel(strep_tb$arm, ref = "Control")
levels(arm_control_ref)
```

`Control` is now first, so it becomes the reference row: the one everything else is compared against. This is the same reference level idea we met with factors in IES Week 7, now doing real work rather than being demonstrated on a copy, and it's worth having used twice before IES Week 11, where setting a sensible reference level is part of fitting a regression model.

We rebuild the table with the releveled factor, and pass it to `riskratio()`:

```{r}
#| label: riskratio-arm
tab_arm_releveled <- table(arm_control_ref, strep_tb$improved)
tab_arm_releveled

riskratio(tab_arm_releveled)
```

`oddsratio()` works the same way. We ask for `method = "wald"`, which uses the same normal-approximation logic behind the ± 1.96 × SE intervals from IES Week 8, and which matches the simple cross-product formula from the hand calculation directly:

```{r}
#| label: oddsratio-arm
oddsratio(tab_arm_releveled, method = "wald")
```

From the table above: 38 of 55 patients improved in the streptomycin arm (risk = 0.69), and 17 of 52 improved in the control arm (risk = 0.33). By hand, risk ratio = 0.69 / 0.33 = 2.11, and odds ratio = (38 × 35) / (17 × 17) = 4.60. Both match what `riskratio()` and `oddsratio()` just reported, with a confidence interval attached to each: a risk ratio of 2.11 (95% CI: 1.38 to 3.24) and an odds ratio of 4.60 (95% CI: 2.04 to 10.39). Neither interval comes close to including 1, the null value for both a risk ratio and an odds ratio, consistent with the chi-squared test above.

`riskratio()` and `oddsratio()` each also report three p-values alongside the estimate: a mid-p exact test, Fisher's exact test, and a chi-squared test. All three agree closely here. The chi-squared one is 0.000166, exactly the p-value `chisq.test(tab_arm, correct = FALSE)` gave us above, to every digit printed. That's the payoff from switching the continuity correction off: the same test, run by two different packages, now gives one number instead of two. Had we left the correction on, `chisq.test()` would have reported 0.0003548, the number IES Week 8's `prop.test()` printed, while `epitools` reported 0.000166, and there'd be no way to tell from the output that the two functions had simply made different default choices.

## `gtsummary::tbl_summary()` with a test attached

`tbl_summary()`, already met in IES Weeks 7 and 8, can attach a test to a stratified table with `add_p()`, piped on afterwards:

```{r}
#| label: tbl-summary-addp
strep_tb |>
  select(arm, improved) |>
  tbl_summary(by = arm) |>
  add_p()
```

This single call reproduces the count, the percentage, and the p-value we built by hand above, in the one-line form we'll reach for once the manual steps are familiar. The p-value matches to every digit because `add_p()` happens to default to the *uncorrected* chi-squared test for a table like this one, the same choice we made by hand with `correct = FALSE`. That's worth checking rather than assuming whenever a shortcut and a manual calculation agree: they agree here because their defaults happen to line up, not because a shortcut is guaranteed to do what you just did by hand.

::: {.callout-note}
## Try it yourself: sample size and significance

Using `slice_sample()` from IES Week 8, take a 10% random subsample of `nhanes` (remember to `set.seed()` first, so the subsample is reproducible) and re-run the `chol ~ gender` t-test from earlier in this chapter on it. Compare the p-value and the confidence interval to the ones from the full sample. Write one sentence on what changed and one sentence on what this means for judging whether a small p-value reflects a large effect or simply a large sample.
:::

## Worked example: sedentary time by gender

Let's bring the independent-samples t-test together on a pairing we haven't used elsewhere in this chapter: sedentary time (`sed`, in hours/day) by gender.

```{r}
#| label: worked-boxplot-sed
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = sed)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Sedentary time (hours/day)")
```

```{r}
#| label: worked-ttest-sed
t.test(sed ~ gender, data = nhanes |> filter(!is.na(gender)))
```

The boxplots overlap substantially, with similar spread in each group, so a t-test is reasonable here too. R prints the difference men minus women once again, so the interval on the page runs from -0.49 to -0.03. Written as it would appear in a results section, with the whole thing reversed: women reported slightly more sedentary time than men (mean difference about 0.26 hours/day, 95% CI: 0.03 to 0.49 hours/day, p = 0.026). The interval is entirely on one side of zero, so the difference is statistically significant, but it's a small effect on a scale of many hours a day.

## Exercises

**Exercise 1** (ILO 1). Perform an independent-samples t-test comparing age (`age`) between men and women. Check the assumption graphically first with a boxplot. Write one sentence reporting the result as it would appear in a results section, and one sentence commenting on whether a statistically significant result here is also a practically important one.

**Exercise 2** (ILO 2). Using `filter()`, restrict `nhanes` to participants whose sedentary time (`sed`) is above the sample median. On this subgroup, run both a paired t-test and a Wilcoxon signed-rank test comparing `met_w` and `met_r`. Write one sentence stating whether your conclusion changes in this more sedentary subgroup, and one sentence explaining why `met_w` and `met_r` are still a paired comparison here, not an independent one, even after filtering.

**Exercise 3** (ILO 3). Construct a contingency table of baseline cavitation (`baseline_cavitation`) against improvement (`improved`) in `strep_tb`. Run a chi-squared test, check the expected counts, and write one sentence interpreting the result.

**Exercise 4** (ILOs 4, 5). Using the same method as the `arm`/`improved` example above, check with `levels()` which category of `gender` is already the reference, then use `relevel()` to set it explicitly (even if it turns out to already be the one you want). Obtain a risk ratio and an odds ratio, each with a 95% confidence interval, for improvement in men compared with women. Reconcile the software output with a hand calculation from the same 2×2 table, and write a two-sentence results statement.

## Writing exercise

Using the `arm`/`improved` result from earlier in this chapter, write two sentences, as they would appear in a results section, for `strep_tb`'s primary outcome. State the proportion improved in each arm, and report the risk ratio or the odds ratio (your choice), its 95% confidence interval, and its p-value, together.

::: {.callout-note collapse="true"}
## Solutions

**Exercise 1**

```{r}
#| label: sol-ex1
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = age)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Age (years)")

t.test(age ~ gender, data = nhanes |> filter(!is.na(gender)))
```

This time R's interval is already positive, and that's informative rather than incidental: the difference is still men minus women, so a positive interval means the first group is the higher one. Men in this sample were on average about 0.7 years older than women (95% CI: 0.4 to 1.1 years, p < 0.001), and no signs need reversing to say so. The difference is statistically significant but tiny relative to a population averaging around 47 years old. With a sample this large, tens of thousands of participants, even a very small true difference will produce a small p-value. Statistical significance here is not the same as a practically important difference, and it doesn't really contradict IES Week 7's impression that age looked similar across the two groups.

**Exercise 2**

```{r}
#| label: sol-ex2
sed_above_median <- nhanes |>
  filter(!is.na(sed), sed > median(sed, na.rm = TRUE))

t.test(sed_above_median$met_w, sed_above_median$met_r, paired = TRUE)
wilcox.test(sed_above_median$met_w, sed_above_median$met_r, paired = TRUE)
```

Both tests still reject the null hypothesis in this more sedentary subgroup, though the mean difference is much smaller than in the whole sample (about 0.49 MET-hours/day here, versus about 3.07 MET-hours/day overall). `met_w` and `met_r` remain two measurements taken on the same person at the same time, just restricted to a different set of people; filtering the rows doesn't change what the columns are measuring, so the comparison is still paired.

**Exercise 3**

```{r}
#| label: sol-ex3
tab_cavitation <- table(strep_tb$baseline_cavitation, strep_tb$improved)
tab_cavitation
chisq.test(tab_cavitation, correct = FALSE)
chisq.test(tab_cavitation, correct = FALSE)$expected
```

All expected counts are comfortably above 5, the smallest being just under 22, so `correct = FALSE` is the right call here for the same reason it was for the `arm` table.

The chi-squared test gives a p-value of about 0.056. That's above the conventional 0.05 threshold, but only just, and it would be a mistake to read much into which side of it the number landed on. The honest reading is the same either way: this is weak evidence of an association between baseline cavitation and improvement. The data lean towards there being one, they don't establish it, and they certainly don't rule it out. A trial of 107 patients simply cannot settle a question this size, and reporting it as "no association" because 0.056 exceeds 0.05 would misrepresent what the analysis found.

**Exercise 4**

```{r}
#| label: sol-ex4
levels(strep_tb$gender)

gender_f_ref <- relevel(strep_tb$gender, ref = "F")
tab_gender <- table(gender_f_ref, strep_tb$improved)
tab_gender

riskratio(tab_gender)
oddsratio(tab_gender, method = "wald")
```

`levels()` shows that `"F"` is already listed first, so `relevel()` here confirms that choice explicitly rather than changing anything. By hand, from the table: 27 of 48 men improved (risk = 0.5625) and 28 of 59 women improved (risk = 0.4746), giving a risk ratio of 0.5625 / 0.4746 = 1.19, and an odds ratio of (27 × 31) / (21 × 28) = 1.42. Both match the software output. The proportion improved was slightly higher in men than in women (risk ratio 1.19, 95% CI: 0.82 to 1.71; odds ratio 1.42, 95% CI: 0.66 to 3.06, chi-squared p = 0.37), but both confidence intervals include the null value of 1, so this trial gives no good evidence of a difference in improvement by gender. `oddsratio()` prints three p-values, as it did for the `arm` table; the chi-squared one is quoted here because that's the one that reconciles with `chisq.test()`.

**Writing exercise (sample answer)**

Among patients receiving streptomycin, 69% (38 of 55) showed radiological improvement at six months, compared with 33% (17 of 52) among those receiving the control treatment. This corresponds to a risk ratio of 2.11 (95% CI: 1.38 to 3.24, p < 0.001), indicating substantially greater improvement with streptomycin than with control in this trial.
:::
