# Correlation and simple linear regression (Week 10) {#sec-10-regression}

This chapter centres on how a continuous outcome varies with a continuous predictor, in contrast with Chapters 4 and 5, where a continuous outcome was compared across the two groups of a binary variable. The tools include scatterplots, Pearson's correlation coefficient, and simple linear regression, specified as Y = α + βX + ε. Simple linear regression also introduces residuals, the intercept, the slope, and R². We discuss and check the model's assumptions. In this chapter we fit that model in R, check it properly, and use it to predict.

## Chapter intended learning outcomes

By the end of this chapter you will be able to:

1. Produce and interpret a scatterplot with a fitted regression line
2. Fit a simple linear regression with `lm()`
3. Generate and interpret standard diagnostic plots
4. Generate predictions and identify when a prediction constitutes extrapolation
5. Extract coefficients and confidence intervals, and report a simple regression result in journal format

## Setup

As in Chapters 3 to 5, we read the data with `read_csv()` and turn the six coded variables into factors against the dictionary.

This chapter adds one new package, `broom`. If `broom` isn't already on your machine, run `install.packages("broom")` once in the console.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(here)
library(broom)

nhanes <- read_csv(here("data", "mph_nhanes_20241107.csv"))
dictionary <- read_csv(here("data", "mph_nhanes_20241107_dictionary.csv"))
```

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
```

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c("Male", "Female")),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c("Mexican American", "Other Hispanic", "White", "Black", "Others")),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c("Less than high school", "High school graduate", "Above high school")),
    smoking = factor(smoking,
      levels = c("Never", "Previous", "Current")),
    phq.c = factor(phq.c,
      levels = c("0 Normal", "1 Mild symptoms", "2 Possible depression"),
      labels = c("Normal", "Mild symptoms", "Possible depression"),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c("No", "Yes"))
  )
```

This is the same recode we introduced in Chapter 3 and repeated in Chapters 4 and 5. See Chapter 3 for the reasoning behind each variable and each choice of `ordered = TRUE`. We repeat it here, unchanged, so this chapter stands on its own too.

## Correlation: Pearson's r and Spearman's rho

The seminar opened with scatterplots and Pearson's correlation coefficient before shifting to regression. Here's the same shift, briefly, in R.

`cor()`, from base R, calculates a correlation coefficient between two continuous variables. Its default method is Pearson's r, which measures the strength of a linear, or straight-line, relationship. We'll use the pairing this chapter's first model uses, `age` and `chol`:

```{r}
#| label: cor-pearson
cor(nhanes$age, nhanes$chol, use = "complete.obs")
```

`use = "complete.obs"` tells `cor()` to drop any participant missing either variable first; without it, a single missing value makes the whole calculation return `NA`. `cor.test()` adds a confidence interval and a p-value testing whether the true correlation is zero, and drops incomplete pairs automatically:

```{r}
#| label: cortest-pearson
cor.test(nhanes$age, nhanes$chol)
```

Pearson's r here is about 0.094 (95% CI 0.084 to 0.104, p < 0.001): a weak but clearly non-zero positive correlation between age and cholesterol. Worth noting once: for one predictor and one outcome, testing whether r differs from zero and testing whether the regression slope differs from zero are the same test, just read on two different scales.

Pearson's r assumes both variables sit on a scale where equal unit of increase/decrease mean the same thing throughout, i.e. an interval or ratio scale. `phq.c` doesn't: "Possible depression" isn't a fixed distance from "Mild symptoms" the way one extra year of age is from another, it's an ordered category, not a measurement. Spearman's rank correlation is the alternative, correlating ranks rather than raw values, so it only needs observations put in order, not a distance measured between them. `phq.c` is already an ordered factor, so `as.numeric()`, from base R, turns it into the numbers 1, 2, and 3 in the order the factor was built, exactly the ranking Spearman's correlation needs:

```{r}
#| label: cor-spearman
phq_numeric <- as.numeric(nhanes$phq.c)
cor.test(nhanes$age, phq_numeric, method = "spearman")
```

Spearman's rho here is about 0.005, p = 0.37: close enough to zero, and far from significant, that age and depression symptom category show no clear monotonic relationship. `cor.test()` doesn't report a confidence interval for Spearman's rho the way it did for Pearson's r above; rho has no equivalent simple formula for one. The warning above, "cannot compute exact p-value with ties," isn't a problem: with tens of thousands of participants spread across only three `phq.c` categories, many necessarily share a rank, so R falls back on an approximate p-value, which is still valid.

Correlation is as far as this chapter takes `phq.c`; the rest follows the seminar's own shift from measuring association to modelling it.

## A scatterplot with a fitted line: cholesterol by age

Chapter 3 plotted total cholesterol (`chol`, in mmol/L) against age (in years) as a plain scatterplot, with no line through it. `geom_smooth()`, from `ggplot2`, adds one. With `method = "lm"`, it fits exactly the simple linear regression this chapter is about, and draws the result:

```{r}
#| label: scatter-age
ggplot(nhanes |> filter(!is.na(age), !is.na(chol)), aes(x = age, y = chol)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Age (years)", y = "Total cholesterol (mmol/L)")
```

Above the plot, `ggplot2` printed `` `geom_smooth()` using formula = 'y ~ x' ``. That's a message, not a warning: nothing has gone wrong, it's just reporting which formula it fitted, since `geom_smooth()` picks one for us if we don't say. Here it chose a straight line in `x`, exactly the simple linear regression we asked for with `method = "lm"`. Worth a glance the first time, since on a plot where we meant something more elaborate it would tell us that we hadn't. It appears above every `geom_smooth()` plot in this chapter and always says the same thing.

The shaded band around the line is a 95% confidence interval for the line itself, the same idea as every other confidence interval in this course, just drawn as a band rather than printed as two numbers. The line slopes gently upward: cholesterol tends to be a little higher in older participants. The cloud of points around it is wide, though. Age clearly doesn't determine cholesterol on its own, and the next section puts a number on exactly how much of the picture the line accounts for.

::: {.callout-note}
## Try it yourself: what the band shows

Re-run the plot above with `geom_smooth(method = "lm", se = FALSE)`. Then try `geom_smooth(method = "lm", level = 0.80)`. Write one sentence on what disappeared in the first change, and one sentence on what changed in the second.
:::

## Fitting the model with `lm()`

`lm()`, from base R, fits a linear model. Its first argument is a formula, in the same `y ~ x` form `t.test()` used in Chapter 4: the variable on the left is what we're modelling, and the variable on the right is what we're modelling it with. We drop the rows where either variable is missing first, the same habit we've used for every model and test so far:

```{r}
#| label: lm-age
age_complete <- nhanes |> filter(!is.na(age), !is.na(chol))
m_age <- lm(chol ~ age, data = age_complete)
summary(m_age)
```

As with `t.test()` in Chapter 4, there's no `$` anywhere in that call. `lm()`'s `data = ` argument tells it to look for `chol` and `age` inside `age_complete`, so it finds them there directly rather than us pulling either one out by hand first. Every model in this course, `lm()` and `glm()` alike, follows this same pattern: name the variables bare in the formula, and point `data = ` at wherever they live.

`summary()` prints a lot at once. The `Coefficients` table has the two numbers that define the fitted line: the `(Intercept)`, the model's predicted cholesterol at age zero, and the slope on `age`, the predicted change in cholesterol for each additional year of age. Each has a standard error, a t-value, and a p-value attached, exactly the reporting standard from Chapter 5. The slope here is small but non-zero: cholesterol tends to be about 0.0056 mmol/L higher for every extra year of age, and the p-value is far below 0.001.

`Multiple R-squared`, R², says what fraction of the variation in cholesterol this model accounts for. Here it's under 0.01: age explains under 1% of the variation in cholesterol between participants. That isn't a contradiction of the significant slope above. A relationship can be real, precisely estimated, and still leave most of the variation in the outcome unexplained. R² and a p-value answer different questions, and neither one substitutes for the other.

`coef()` and `confint()`, both base R, pull out just the coefficients and just their confidence intervals, without the rest of the printout:

```{r}
#| label: coef-confint-age
coef(m_age)
confint(m_age)
```

`broom::tidy()` puts the same information into a tidy data frame, one row per coefficient, which is easier to work with in code than `summary()`'s printed output. Adding `conf.int = TRUE` includes the confidence interval as two extra columns:

```{r}
#| label: tidy-age
tidy(m_age, conf.int = TRUE)
```

Written as it would appear in a results section: total cholesterol was on average 0.0056 mmol/L higher for each additional year of age (95% CI: 0.0050 to 0.0061 mmol/L, p < 0.001), although age accounted for less than 1% of the variation in cholesterol between participants.

## Checking assumptions: diagnostic plots

A fitted line is only trustworthy if the model's assumptions hold: linearity (a straight line is the right shape), constant variance (the scatter around the line doesn't change as we move along it), and normally distributed residuals (the vertical distances between the points and the line). A fourth assumption, independence between observations, isn't something a plot from one fitted model can check. It's a property of how the data were collected, and NHANES's sampling design is what would need inspecting for that, not this model's output.

`broom::augment()` adds each observation's fitted value and residual back onto the original data, in columns named `.fitted` and `.resid`:

```{r}
#| label: augment-age
age_aug <- augment(m_age)
age_aug
```

Plotting residuals against fitted values checks linearity and constant variance together. If the model is a good fit, the points should scatter evenly above and below zero, with no obvious curve and no obvious change in spread from left to right. `geom_hline()`, from `ggplot2`, draws a single horizontal line across the plot at whatever height `yintercept` names, and we put one at zero to make "above and below" easy to judge by eye:

```{r}
#| label: resid-fitted-age
ggplot(age_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted value (mmol/L)", y = "Residual (mmol/L)")
```

Most points sit in a fairly even band around zero, consistent with linearity and roughly constant variance. A small number of participants with very high cholesterol readings sit well above the line, though, visible here as points a long way above zero. Those same participants are worth keeping in mind for the QQ plot, which checks whether the residuals are approximately normally distributed. `.std.resid`, also added by `augment()`, standardises each residual to the same scale a normal distribution would use. `stat_qq()` and `stat_qq_line()`, both from `ggplot2`, plot the standardised residuals against what a perfect normal distribution would produce, with a reference line for comparison:

```{r}
#| label: qq-age
ggplot(age_aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Standardised residuals")
```

Points that fall on the line are consistent with normally distributed residuals. Here, the middle of the distribution sits close to the line, but the upper tail pulls away from it. That's the same small group of participants with unusually high cholesterol showing up again: the residuals are a little more heavy-tailed than a normal distribution would predict. With a sample this large, the model's estimates are still reliable despite this. With a much smaller sample, tail behaviour like this would be more of a concern.

## Prediction and extrapolation from the model

`predict()`, from base R, uses a fitted model to generate a predicted value for a new set of inputs. Those inputs need to arrive as a data frame, with a column named to match each predictor in the model. `data.frame()`, also from base R, builds one directly from the values we supply, rather than reading one in from a file:

```{r}
#| label: predict-age
predict(m_age, newdata = data.frame(age = c(45, 150)), interval = "confidence")
```

`interval = "confidence"` adds a confidence interval around each prediction. This confidence interval describes the precision of the predicted **mean** value, not the precision for an **individual**. The precision around an individual prediction is called a **prediction interval**, and it's always wider than the confidence interval around the mean, since predicting one person's value is less certain than predicting the average across many. A prediction interval can be requested by stating `interval = "prediction"` instead.

The first row, age 45, sits comfortably inside the range of ages actually observed in this sample (18 to 85), so this prediction is supported by data the model was fitted on. The second row, age 150, is not. Nobody in this dataset, or in any verified human population, is 150 years old. The model has no information about what happens at that age, and fitting a straight line through the data we do have says nothing reliable about what lies far outside it. The predicted value it returns looks like an ordinary number, which is exactly what makes extrapolation dangerous: a plausible-looking output doesn't mean the input that produced it was reasonable. Always check a prediction's input against the range the model was actually fitted on before trusting its output.

## Worked example: cholesterol by sedentary time

Let's bring the whole workflow together on another research question: is sedentary time (`sed`, in hours/day) associated with total cholesterol (`chol`)?

```{r}
#| label: worked-scatter-sed
ggplot(nhanes |> filter(!is.na(sed), !is.na(chol)), aes(x = sed, y = chol)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Sedentary time (hours/day)", y = "Total cholesterol (mmol/L)")
```

The fitted line is nearly flat, and the confidence band around it is wide enough to include a flat, horizontal line, that is, no relationship at all. Fitting the model directly confirms what the plot suggests:

```{r}
#| label: worked-lm-sed
sed_complete <- nhanes |> filter(!is.na(sed), !is.na(chol))
m_sed <- lm(chol ~ sed, data = sed_complete)
summary(m_sed)
confint(m_sed)
```

The slope is small and negative, about -0.0010 mmol/L for each additional hour/day of sedentary time, and the p-value is 0.073: above the conventional 0.05 threshold. The 95% confidence interval for the slope, -0.0021 to 0.0001 mmol/L, includes zero. R² is essentially zero. On this evidence alone, sedentary time and cholesterol show no clear linear relationship in this sample.

The diagnostic plots are worth checking anyway, since a weak result could be caused by violated assumptions:

```{r}
#| label: worked-augment-sed
sed_aug <- augment(m_sed)

ggplot(sed_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted value (mmol/L)", y = "Residual (mmol/L)")

ggplot(sed_aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Standardised residuals")
```

The same pattern from the age model shows up again, more visibly: a handful of participants sit far from the fitted line, and `sed` itself carries a long right tail, including some participants recording well over 24 hours of sedentary time a day, which cannot be literally true and points to some noise in how this variable was recorded or reported. None of this overturns the conclusion above. It's a reason to hold the result loosely rather than to distrust the direction of the estimate.

Written as it would appear in a results section: sedentary time was not associated with total cholesterol in this sample (slope -0.0010 mmol/L per hour/day, 95% CI: -0.0021 to 0.0001 mmol/L, p = 0.073). But it would be a mistake to conclude from this alone that sedentary time has no bearing on cholesterol. This model adjusts for nothing else: age, gender, and other characteristics could still be masking a real association. Chapter 7 revisits this same pairing with those factors added back in.

## Exercises

**Exercise 1** (ILOs 1, 2, 5). Using `vpa_r`, vigorous physical activity outside work (hours/day), produce a scatterplot of `chol` against `vpa_r` with a fitted regression line. Fit the model with `lm()`, and extract its coefficients and their 95% confidence intervals with `confint()` and `broom::tidy()`. Write one sentence reporting the slope as it would appear in a results section.

**Exercise 2** (ILO 3). For the model fitted in Exercise 1, produce a residuals-vs-fitted plot and a QQ plot. Write one sentence describing what each plot shows, and one sentence commenting on anything the plots reveal about the shape of `vpa_r` itself.

**Exercise 3** (ILO 4). Using the model fitted in Exercise 1, predict total cholesterol at `vpa_r` = 1 hour/day and at `vpa_r` = 24 hours/day. Write one sentence stating which of the two predictions can be trusted, and one sentence explaining why the other cannot, referring to the range of `vpa_r` actually observed in the data.

## Writing exercise

Using the `chol ~ sed` result from the worked example, write a short results paragraph of two to three sentences, as it would appear in a report. State the estimated slope, its 95% confidence interval, and its p-value, and say plainly what this evidence does and does not support. Avoid two common mistakes: don't describe a non-significant result as proof there is no relationship at all, and don't write around the p-value without stating it.

::: {.callout-note collapse="true"}
## Solutions

**Exercise 1**

```{r}
#| label: sol-ex1
vpa_r_complete <- nhanes |> filter(!is.na(vpa_r), !is.na(chol))

ggplot(vpa_r_complete, aes(x = vpa_r, y = chol)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Vigorous activity outside work (hours/day)", y = "Total cholesterol (mmol/L)")

m_vpa_r <- lm(chol ~ vpa_r, data = vpa_r_complete)
summary(m_vpa_r)
confint(m_vpa_r)
tidy(m_vpa_r, conf.int = TRUE)
```

Total cholesterol was about 0.170 mmol/L lower for each additional hour/day of vigorous activity outside work (95% CI: -0.201 to -0.139 mmol/L, p < 0.001).

**Exercise 2**

```{r}
#| label: sol-ex2
vpa_r_aug <- augment(m_vpa_r)

ggplot(vpa_r_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted value (mmol/L)", y = "Residual (mmol/L)")

ggplot(vpa_r_aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Standardised residuals")
```

The residuals-vs-fitted plot checks linearity and constant variance; the QQ plot checks whether the residuals are approximately normally distributed. Most points in the residual plot fall into a small number of vertical strips rather than scattering freely, because `vpa_r` itself only takes a small number of distinct values: most participants report exactly zero hours/day of vigorous activity outside work, so their fitted value and residual pattern is shared rather than spread out. This is a feature of the predictor's own distribution, not a sign that the model is wrongly specified.

**Exercise 3**

```{r}
#| label: sol-ex3
predict(m_vpa_r, newdata = data.frame(vpa_r = c(1, 24)), interval = "confidence")
```

The prediction at `vpa_r` = 1 hour/day (about 4.80 mmol/L) can be trusted, because an hour of vigorous activity outside work a day is well within the range observed in this sample (0 to 8 hours/day). The prediction at `vpa_r` = 24 hours/day (about 0.90 mmol/L) cannot: 24 hours/day of vigorous activity is not physically possible to sustain, it is far outside the observed range, and the resulting prediction, a cholesterol level barely above zero, is not a value any living person could have. It is a consequence of extending the fitted line somewhere the data give it no support, not a real finding.

**Writing exercise (sample answer)**

Sedentary time was not significantly associated with total cholesterol in this sample (slope -0.0010 mmol/L per hour/day, 95% CI: -0.0021 to 0.0001 mmol/L, p = 0.073). The confidence interval includes zero, so these data are consistent with no true association. This model adjusts for nothing else, so it cannot say whether age, gender, or other characteristics are masking or accounting for this result.
:::
